Going from coarse landings data to fine scale species distribution

### 2022-04-26 Version_M2_plusieurszones/draft/paper_3/paper_3.Rmd
intro

---
title             : "Going from coarse landings data to fine scale species distribution"

@@

References for change of support issues in integrated modelling:

https://esajournals.onlinelibrary.wiley.com/doi/epdf/10.1002/ecy.2710

https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12793

https://www.sciencedirect.com/science/article/pii/S0169534719302551#b0010

https://www.sciencedirect.com/science/article/pii/S0006320721001993

>>>>

---
title             : "Downscaling coarse observations to predict continuous wild species spatial distribution"

@@

# Introduction

**Attention:** L'introduction n'est pas finie

**Context**

Spatial ecological data are becoming more and more accessible every days thanks to the huge effort from the scientific community to provide information for ecology, evolution and conservation [@nathan2022big;@hampton2013big;@gremillet2022big]. These data are crucial to face the current challenges related to both large and small scales ecological issues (e.g. tracking climate change [@maureaud_are_2020], tracking animal movement [@nathan2022big], mapping species distribution [@isaac2020data]). In particular, a key point of these studies lies in understanding what are the drivers of the ecological processes. For instance, in species distributions modelling, one will try to find the main covariates (e.g. substrat, temperature, depth/altitude) that shape species distribution [@guisan_predictive_2000;@planqueUnderstandingWhatControls2011]. Combining these new datasources and more standard datasources is expected to improve our understanding of these processes and to fill the gaps in the blind spots of ecology. However, these datasources are highly heterogeneous in size, type and sampling design making there combination a methodological challenge [@pacifici2017integrating;@fletcher_practical_2019;@miller2019recent;@renner2019combining;@isaac2020data]. For instance, combining scientific standardized data and citizen science data has been an important question in the last decade as they prooved to be highly complementary. The first one benefit from standardized protocol, controlled sapling plan and is usually considered as high-quality data. The second one provide a larger amount of data with lower cost, but is also caracterized by misreporting or opportunistic sampling and then by potential sampling bias [@botella2021jointly;@steen2021spatial].

Another less frequently mentionned data are declarations data (we refer to declaration data as the mandatory data that must be reported by some agent as a legal requirement to proceed with his activity). As they are mandatory, declarations data are usually very large datasets and thus they may be very informative of species spatial ecology. A common example are commercial catch declarations data in fisheries science [@hintzenVMStoolsOpensourceSoftware2012] or harvest data in hunting [@bauder2021mismatched;@gilbert2021integrating]. These data are most often declared at the scale of coarse spatial units while scientific survey (and commonly citizen science data) are usually reported with their exact locations. Furthermore, these administrative units do not generaly have a resolution that is relevant for ecological analysis and conservation [@pacifici2019resolving].

Consequently (in ecological applications but in other fields too), to bypass the difference of resolution it is standard to apply rough hypothesis through geoprocessing technics based on proportional allocation and centroid smoothing [@gotway2007geostatistical;@young2007linking;@hintzenVMStoolsOpensourceSoftware2012]. However, these methods do not provide a measure of uncertainty related to the processing of the data and can lead to bias not only in estimates of the mean and the variance, but also in any statistic that would be derived from these estimates. Overall, consequences of such procedures are hard to predict and can result in strong and deleterious biases [@gotway2007geostatistical;@pacifici2019resolving]. Developing statistical methods to properly infer spatial ecological process at a fine scale from coarse declarations data and integrate these with higher resolution data is then a challenge that could give access to huge amount of information.

**The COS issue**

Reconciling spatial (and eventually temporal) scales properly when different set of observations do not have the same resolution is a well known issue in geography, ecology, agriculture, geology and statistics [@gotway2002combining]. Inferring values of a variable at points or blocks different from those that have been observed is commonly refered as the *change of support problem* (COS) in the statistical literature [@gotway2002combining and @gelfand2010misaligned]. COS issues can be classified within 2 categories - that are not necessarily mutually exclusive - [@pacifici2019resolving]:

- (1) 'spatial and temporal misalignement' (when for example some multivariate data, where not all variables have been observed over all locations - see for example @finley2014bayesian).

- (2) 'modifiable areal unit' problem (MAUP) when data are aggregated over increasingly larger geographic scales (e.g. collected at point level but regrouped/declared at coarse level - see for instance @gilbert2021integrating). Our case of application clearly refers to MAUP. Note that within this category, another category exist and is commonly refered to the 'ecological fallacy'. This occures when the individual response to a covariate differs from the response estimated from grouping individuals. This means that conclusions from a fine-resolution analysis can differ from an analysis at a coarser scale based on the aggregation of the fine-resolution data.

Since 2000, several studies have described how could be overcome the different COS issues; @mugglin2000fully, @gelfand2001change, @gotway2007geostatistical, @wikle2005combining proposed generic approaches (and extensions of these approaches - @kim2016change) for adressing COS in spatial or spatio-temporal context. In health analysis @young2007linking proposed to compare some standard (and rough) way to connect data that do not have the same spatial resolution with a method properly accounting for COS. @berrocal2010bivariate and @berrocal2010spatio proposed a spatio-temporal method for fusing several air pollution data: one from coarse resolution but with full spatial coverage and another recorded at point level, with sparse distribution but where records almost corresponds to the true value of the process. In climate science, @reich2014spectral and @parker2015multiresolution proposed a spectral statistical approach to downscale information from large scale model to lower scales. In the field of ecology, few recent studies have tackled such issues: @finley2014bayesian provided a framework for integrating spatially misaligned data, @hefley2017bias proposed a solution based on COS to account for location error in presence-only data, @pacifici2019resolving introduced a framework for integrating datasources of different resolution to map species distribution. Applying similar ideas, @gilbert2021integrating integrated harvest data (aggregated data) and camera trap (punctual data) to map several wildlife species in Wisconsin.

**The point that we adress in this paper**

Still, applications in ecology remains sparse in regards to the huge amount of data that is available. Besides, one of the limiting point consist in the kind of observational data that can be fitted to the existing COS framework. Indeed, the framework that were developped so far and their related applications mainly limited their scope to relatively simple observational data: count data were modelled through Poisson processes [@mugglin2000fully;@gotway2007geostatistical;@pacifici2019resolving;@gilbert2021integrating] and continous data were modelled through Gaussian or Gamma distributions [@gelfand2001change;@wikle2005combining;@berrocal2010spatio]. However, ecological data do not always consist of observation that can be modelled with standard probability distributions. For instance, in frequent cases data may be zero-inflated and positive-continous data. Several studies have developed models to handle properly such punctual data in a computationaly efficient way [@lecomte2013compound;@thorson2018three]. However, these may complicate a bit the way COS is tackled when dealing with an aggregation of such complex data as their convolution may not be as simple as Poisson or Gaussian ones.

In this paper, we aim at illustrating how to deal with change of support in ecological applications when the observational data is complex (e.g. zero-inflated and positive continous). We develop our approach on an existing framework developped by @alglave_combining_2022 in the field of marine and fisheries ecology. The framework aims at predicting the spatial distribution of fish species based on 2 datasets: scientific survey data and commercial catch declarations data. Commercial catch declarations are declared at the level of ICES rectangles (resolution of 0.5 x 1) while scientific data benefit from exact location records. Usually in standard processing, declarations data are reallocated uniformly over there GPS fishing positions (available through Vessel Monitoring Satellites - VMS) in order to improve its spatial resolution.

In the following, we first describe the original model integrating both datasources (part 1). We then propose a generic statistical solution that allow to properly tackle the change of support issue and adapt it to our specific case (part 2).

Finally, we compare this method to the current geoprocessing method in 3 steps:

- A first simulation-estimation step at the scale of a single statistical rectangle where a the biomass distribution is simplified in order to explore the base properties of our framework, a more realistic simulation-estimation step (part 3).

- A second simulation-estimation step at the scale of several statistical rectangles. The latent field is complexified and the set of simulation is parameterized so as to get closer to a realistic case study (part 4).

- A final step where we compare the 2 methods on a real case study (common sole in the Bay of Biscay) and illustrate the main results that are consitent with simulations-estimations (part 5).
